Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-26 manu-draft.qmd
working on methods

Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed includes another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats. The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed. Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.

Habitats generally described as 'subtidal' include hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] are classified as emergent tidal wetlands. For the purposes of this planning effort, supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* As discussed below, uplands are sub-divided into coastal and non-coastal uplands, based on location relative to the 5-foot contour.

@@

Recommended targets were created stepwise with geospatial analyses: 1) quantifying habitat status and historical trends; 2) synthesizing historic habitat restoration efforts; and 3) identifying and defining remaining 'opportunities' for restoration that integrated results from the first two analyses. Final development of restoration targets and goals integrated strata and opportunity area information with an analysis of past restoration implementation to use previous completed projects as a guide to a feasible amount of restoration that could be conducted by partners. This approach will ensure that the estuarine-dependent species and faunal guilds throughout the watershed, as defined in the original approach, will continue to be supported.

@@

For the majority of subtidal, intertidal and supratidal habitats, primary data derived from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD) were utilized. However, to address data gaps for some habitats, results from special studies were integrated with the primary data sources. These included hard bottom [@Kaufman17; @CSA19], dredged holes [@Griffen05; @Raulerson19] and oyster habitat [@OKeife06]*.*

The source data used to estimate the most current coverage of seagrasses, tidal flats, and oysters was the *Seagrass in 2018* geospatial database [@SWFWMD19]. The bi-annual seagrass monitoring program was initiated in 1988 under SWFWMD's Surface Water Improvement and Management program [@Sherwood17; @Tomasko20]*.* SWFWMD has estimated oyster bed coverage as part of this program since 2014.

The source data used to estimate and map trends in development, emergent tidal wetlands, freshwater wetlands, and native upland habitats was the SWFWMD Land Use Land Cover (LULC) series geospatial database [@SWFWMD18]. This comprehensive database classifies the land use and cover types (natural and developed) pursuant to the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often very difficult, it is possible to accurately distinguish forested wetlands from non-forested wetlands. Therefore, for this analysis, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications. Similarly, within the target- and goal-setting exercise, uplands were combined in one classification. These classifications were reported for each mapping exercise conducted every 2-3 years from the start of the program 1990 through 2017, the most recent year with available data.

To quantify the extent of tidal creeks, GIS data [@JEI16; @JEI20] was clipped to the Florida Department of Environmental Protection stream segments that were classified as estuarine.

@@

We quantified past restoration efforts in each of the major habitat types to guide decisions on future targets and goals. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database. Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the TBEP. Primary data collected included project name, year, description, lead partner, size (area or length), and latitude and longitude. Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and documenting the knowledge of local professionals. Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.

@@

@fig-gisworkflow

@@

Opportunity areas, defined here as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets, were also analyzed. The definition and mapping of opportunity areas is necessary to quantify the "restoration potential" for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions. The most appropriate opportunity areas are generally not developed and located on existing public lands or areas identified for acquisition.

The Land Use Land Cover 2017 geospatial database (Table 1) was used as the baseline for cataloguing existing and opportunity areas for intertidal and supratidal habitats in the Tampa Bay watershed. All FLUCCS classification codes were placed into one of three categories. First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands). Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography. Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.

>>>>

Tampa Bay is a large open water estuary (surface area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats. The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties of the watershed [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed. Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the reaminder of the watershed).

Subtidal included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] were classified as emergent tidal wetlands. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.*  Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).

@@

Opportunity areas for restoration and recommended coverage targets for habitat types were identifed by integrating multiple datasets available for the region.  Habitat status and historical trends were first quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.  Historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.  Finally, remaining opportunities for restoration that integrated results from the first two analyses were identified based on current status of habitat coverage in the watershed.  This included combining current coverages with existing protected areas and considering their location within the three strata (sub-, inter-, supratidal) and areas anticipated to be impacted by sea-level rise and development near coastal areas.  The final results included spatially-explicit maps that identified opportunity areas and restoration potential for remaining native habitats in the watershed.  Short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats were informed by current status, past trends, and historical restoration efforts in the region.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19]. Historical datasets for subtidal habitats begin in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) geospatial database [@SWFWMD18]. Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands from non-forested wetlands can be distinguished with these data. Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications. Native upland habitats were also combined in one classification. Historical estimates for these habitat coverages were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.

To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates. These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.

Current and historical coverage estimates were obtained for all habitat types, including estimates of developed lands, for each year of available data spanning approximately three decades at two-year time steps.  In addition, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types.  This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).  The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years.

@@

Past restoration projects were quantified for each of the major habitat types to inform expectations for setting short-term goals and long-term targets. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database. Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, lead partner, size (area or length), and latitude and longitude. Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals. Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.

@@

Spatially-explicit estimates of the remaining opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.  The two main processes included binning existing datasets into relevant categories and overlaying multiple datasets to identify restoration opportunities.  Opportunity areas were defined as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets. The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related both to land use change and sea-level rise.

The Land Use Land Cover 2017 dataset from SWFWMD was used as the baseline for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata.

 All FLUCCS classification codes were placed into one of three categories. First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands). Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography. Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.

@@
